Estimating demographic bias on tests of children's early vocabulary

### 2022-01-17 paper/cogsci.Rmd
writing

In the multigroup Rasch model, an item's difficulty is allowed to vary depending on which group an individual child belongs to. 
For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
We fitted a multigroup Rasch model for each demographic dimension (sex, socioeconomic status, and ethnicity).
To contextualize these results, we also first fitted a baseline Rasch model

@@

Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects in the expected directions, as can also be seen in the figure. 
Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$.
High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
We will re-examine these demographic regressions after trimming items showing extreme DIF, which is what we turn to now.

@@

Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
The full GLIMMERs, with all 680 CDI:WS items, are available on OSF, but were too large to include here.
Nonetheless, it is important to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.
For the SES and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].

@@

mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))

mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.13 max: 1.07
mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56

# median sex difference: 
# median(mm_sex$d_diff) # 0.28

# median SES difference: 
# median(mm_ses$d_diff) # 0.009 - very small!

# median race difference:
# median(mm_eth$d_diff) # 0.40 - large!

# report # of items above and below 0 for each demographic?

all_extreme <- rbind(subset(mm_sex, extreme),
                     subset(mm_ses, extreme),
                     subset(mm_eth, extreme)) 
# length(unique(all_extreme$definition)) # 76 - so 19 items extreme on >1 dimension
# sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
```

@@

## Demographic effects after trimming

Having identified 76 CDI items that function differently across demographic groups -- including 19 that

>>>>

In the multigroup Rasch model, an item's difficulty is allowed to vary depending on which group an individual child belongs to. 
For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
In the quest to identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, socioeconomic status, and ethnicity).
To contextualize these results, we first fitted a baseline Rasch model to the entire dataset, without demographic information.

@@

Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects in the expected directions, as can also be seen in the figure. 
Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$).
High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
We will re-examine these demographic regressions after trimming items showing extreme DIF, which is what we turn to now.

@@

Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
The full GLIMMERs, with all 680 CDI:WS items, are available on OSF, but were too large to include here.
Nonetheless, it is important to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.
For the SES and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Nonetheless, we describe the distributions of item-level group difficulty differences, and examine overlap of the extrema.

@@

mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("low","high"))
mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))

mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56

all_extreme <- rbind(subset(mm_sex, extreme),
                     subset(mm_ses, extreme),
                     subset(mm_eth, extreme)) 
# length(unique(all_extreme$definition)) 
# sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom

# overlap between extreme SES and race items:
#intersect(subset(mm_ses, extreme)$definition, subset(mm_eth, extreme)$definition)
```

For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
The fact that the bulk of this distribution favors females seems to confirm that the female language advantage is pervasive across the CDI:WS, and is thus likely to be a real advantage.
For SES, the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-SES than low-SES children. 
With the mean and median difficulty differences close to 0, and roughly half of the words favoring each SES group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to SES.
For race, the median difficulty difference (nonwhite-white) was `r round(median(mm_eth$d_diff),2)` (M=`r round(mean(mm_eth$d_diff),2)`, sd=`r round(sd(mm_eth$d_diff),2)`), with `r length(which(mm_eth$d_diff>0))`/680 items being easier for white than non-white children, revealing a fairly pervasive advantage for white children on CDI items.

Finally, for each distribution we looked at extreme values, flagging items with difficulty differences that were outside 2 standard deviations.
This identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males. <!-- thus, removing these extrema would actually lower male children's vocabulary scores..] -->
For SES, 39 extrema were identified, only 13 of which were easier for low-SES children.
For race, 31 extrema were identified, only 11 of which were easier for non-white children.
Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39 extreme items in the SES model favored high-SES children, and 20/31 items in the race model favored white children.
17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
In our final analysis, we explore the effect of pruning the 15 items that were extreme in both SES and race models.

```{r items-to-remove}
prune_defs <- c("tractor", "vroom", "grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy*", "pet's name", "up", "so") # also "choo choo" and "give me five!" ?
prune_cols <- which(!is.element(names(d_mat), prune_defs))
#names(d_mat)[which(is.element(names(d_mat), prune_defs))]
d_mat_pruned <- d_mat[,prune_cols]
```

```{r fit-pruned-model, eval=F}
m0 <- mirt(d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(-data_id), 1, "Rasch", 
           technical=list(NCYCLES=1000))

coefs_1pl <- as_tibble(coef(m0, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(m0, simplify = TRUE)$items))
fscores_1pl <- tibble(data_id = d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(data_id), 
                             ability = fscores(m0, method = "MAP")[,1])

save(file = here("data/eng_Ws_prod_mod_1pl_pruned.Rds"), "m0", "coefs_1pl", "fscores_1pl")
```

@@

## Demographic effects after pruning

```{r load-pruned}
load(here("data/eng_Ws_prod_mod_1pl_pruned.Rds"))

# join with d_demo, do the demo regressions again..
```

We identified 76 CDI items that function quite differently across demographic groups, and chose to prune the 15 items with large difficulty differences for both SES and race.
